EDIpack: A generic and interoperable Lanczos-based solver for quantum impurity problems

### 2025-05-28 01_intro.tex
Huge iteration
- Igor passed through a major part of the manuscript
- Samuele edited some parts of the e-ph section (response to Massimo)
- Adriano added author contributions
- Gabriele answered some comments by Igor and Lorenzo on the iRDM

--------

Co-authored-by: Igor Krivenko <iskrivenko@proton.me>
Co-authored-by: Lorenzo Crippa <crippa.lorenzo@gmail.com>
Co-authored-by: Samuele Giuli <samuele.giuli@gmail.com>
Co-authored-by: Adriano Amaricci <adriano.amaricci@gmail.com>

\section{Introduction and Motivation}\label{SecIntro}
\fixme{A tentative Introduction scheme}

Quantum impurity models play a central role in the study of strongly correlated electron systems, capturing the coupling of a small number of localized degrees of freedom with extended non-interacting environment.   
Originally introduced to describe diluted magnetic impurities in metallic hosts, single-impurity models have led to the full understanding of the Kondo problem, thereby progressively becoming powerful tools for exploring a wide range of quantum many-body phenomena. More recently, their importance has extended beyond the realm of condensed matter, influencing understanding of other areas such as quantum information or ultra-cold atomic systems.
%Asymptotic freedom

At the heart of quantum impurity models is the interaction between a localized ``impurity"  and a surrounding bath of itinerant particles, usually represented as a conduction band of electrons. In this context, the definition of the impurity can taken as rather general, in order to contain lorbital and spin local degrees of freedom as well as a finite number of lattice sites. %MC eliminated inequivalent to avoid confusion with the last point
This setup captures many essential aspects of strong correlation including screening, decoherence and entanglement in a computationally tractable way. 
Most notably, quantum impurity problems lie at the core of dynamical mean-field theory (DMFT) where the complex behavior of the lattice systems is reduced to an effective self-consistent quantum impurity problem.  
By accounting for local quantum fluctuations, DMFT can successfully describe the most important physical features of correlated materials, including Mott insulators, heavy fermion compounds, and high-temperature superconductors.

Accurately and efficiently solving a quantum impurity problem and computing the Green's functions which are central in the DMFT framework, remains a
critical challenge, particularly for multi-orbital systems, systems
at low temperatures or endowed with lower symmetry. During many years of development, different advanced numerical techniques have been developed together with cutting edge approaches aimed to reduce computational load.

%EDIPACK
In this work we introduce  \NAME{}: a flexible, high-performance numerical library designed to address these challenges, providing an exact diagonalization (ED) solver for generic Quantum Impurity Problems. 
The library builds on the foundations of
its predecessor, featuring massively parallel Lanczos-based algorithms while including several significant extensions and  improvements. The library now supports, within a unified framework, both zero and finite temperature calculations for a wide set of impurity models, making it capable of addressing a large set of problems, from spin-orbit coupling effect in quantum materials to multi-orbital superconductivity. It also provides direct
access to well-approximated %MC why? I would say numerically exact 
dynamical correlation functions across the
entire complex plane, thus enabling calculations of spectral functions, local susceptibilities, and other response properties.

@@

%MC this last point is intrinsically DMFT (or embedding) and it has no connection with impurity physics
This new version also includes support for systems with non-equivalent impurities , a critical feature to study hetero-structures, super-lattices, and disordered systems, where the local environment varies significantly across different sites. This capability makes it 
possible to explore a wide range of inhomogeneous systems, including
oxide interfaces or disordered systems, where
translational symmetry is broken.

%\paragraph{Interoperability}\\
With the growing importance of  quantum impurity models in the study of correlated materials,
standardizazion and interoperability between solution software suites has
become a critical necessity.
\NAME has been expressly designed following this guiding principle. 
While the library itself is implemented using modern Fortran constructs, it also provides an extensive set of C-bindings that enable integration with other programming languages,
be they compiled (C, C++), JIT-compiled (Julia) or interpreted (Python). 
Especially notable in this sense is the Python API, EDIpy which, beyond providing a 
high-level interface for defining and solving impurity problems, also allows integration with established scientific libraries, such as NumPy and SciPy, facilitating data analysis and
rapid prototyping.

A key development feature in \NAME, unlocked by the introduction of Python and C/C++ API, 
is the direct interfaces to the TRIQS (Toolbox for Research on Interacting Quantum Systems) and
W2Dynamics software suites, which are two of the most important and used computational tools in
the field. \NAME seamlessly %MC ?
slots in as an alternative solver to the native Quantum Monte Carlo 
implementations, greatly extending the parameter (and especially temperature) range reachable by 
the simulation while minimizing the coding requirements for the end user.
At the same time, this integration provides \NAME users with access to the
extensive data handling and analysis tools in the TRIQS ecosystem. 
Overall, interoperability extends the cumulative reach and trustworthiness of all these platforms,
enhancing reproducibility and cross-validation of numerical results, with the ultimate aim of a
chieving standardization for quantum impurity solvers along the guiding principles of initiatives such as FAIRmat.

%\paragraph{closure}
In this paper, we present the mathematical framework, implementation
details, and key features of \NAME, along with several illustrative
applications. We aim to demonstrate how this flexible,
high-performance solver can be used to tackle a wide range of
challenging impurity problems. 
The rest of this work has the following structure. In \secu{SecInstall} we 
briefly illustrate the overall structure of \NAME together with its dependencies, configuration
and installation procedures. In the following  \secu{SecEDIpack} we present the quantum impurity
problem and review the software implementation in full details. In the following 
\secu{SecInterop} we thoroughly discuss the C-binding interface which is at the heart of the 
interoperability capabilities of \NAME. In this section we also discuss 4 interface layers: i) 
the Python API, ii) the TRIQS interface, iii) the W2Dynamics interface and iv) a basic Julia API.
Finally, in \secu{SecExamples} we discuss the functionalities and the capabilities of \NAME and 
all the discussed extensions with a series of illustrative examples.

>>>>

\section{Introduction and Motivation}\label{SecIntro}
Quantum impurity models play a central role in the study of strongly correlated electron systems, capturing the coupling of a small number of localized degrees of freedom with an extended non-interacting environment.   
Originally introduced to describe diluted magnetic impurities in metallic hosts, single-impurity models have led to the full understanding of the Kondo problem, and beyond that they have progressively become powerful tools for exploring a wide range of quantum many-body phenomena. More recently, their importance has extended beyond the realm of condensed matter, influencing understanding of other areas such as quantum information or ultra-cold atomic systems.
%Asymptotic freedom

At the heart of quantum impurity models is the interaction between a localized ``impurity''  and a surrounding bath of itinerant particles, usually represented as a conduction band of electrons. In this context, the definition of the impurity can be taken as rather general, in order to contain orbital and spin local degrees of freedom as well as a finite number of lattice sites. %MC eliminated inequivalent to avoid confusion with the last point
This setup captures many essential aspects of strong correlation including screening, decoherence and entanglement in a computationally tractable way. 
Most notably, quantum impurity problems lie at the core of dynamical mean-field theory (DMFT) where the complex behavior of the lattice systems is reduced to an effective self-consistent quantum impurity problem \cite{Georges1996RMP}.  
By accounting for local quantum fluctuations, DMFT can successfully describe the most important physical features of correlated materials, including Mott insulators, heavy fermion compounds, and high-temperature superconductors.

Accurately and efficiently solving a quantum impurity problem and computing the Green's functions, which are central in the DMFT framework, remains a
critical challenge, particularly for multi-orbital systems, systems
at low temperatures or endowed with lower symmetry. During many years of development, different advanced numerical techniques have been developed together with cutting edge approaches aimed to reduce computational load.

%EDIPACK
In this work we present the renewed and improved \NAME{}, version number \textcolor{red}{5.0}: a flexible, high-performance numerical library designed to address these challenges, providing an exact diagonalization (ED) solver for generic Quantum Impurity Problems. 
The library builds on the foundations of
its predecessor~\cite{Amaricci2022}, featuring massively parallel Lanczos-based algorithms while including several significant extensions and improvements. The library now supports, within a unified framework, both zero and finite temperature calculations for a wide set of impurity models, making it capable of addressing a large set of problems, from spin-orbit coupling effects in quantum materials to multi-orbital superconductivity. It also provides direct
access to numerically exact
dynamical correlation functions across the
entire frequency complex plane, thus enabling calculations of spectral functions, local susceptibilities, and other response properties.

@@

%MC this last point is intrinsically DMFT (or embedding) and it has no connection with impurity physics
This new version also includes support for systems with non-equivalent impurities. This is a critical feature in DMFT or, more generally, in quantum embedding methods, to study hetero-structures, super-lattices, and disordered systems, where the local environment varies significantly across different sites. This capability makes it 
possible to explore a wide range of inhomogeneous systems, including
oxide interfaces or disordered systems, where
translational symmetry is broken.

%\paragraph{Interoperability}\\
With the growing importance of  quantum impurity models in the study of correlated materials,
standardization and interoperability between solution software suites has
become a critical necessity.
\NAME has been expressly designed following this guiding principle. 
While the library itself is implemented using modern Fortran constructs, it also provides an extensive set of C-bindings that enable integration with other programming languages,
be they compiled (C, C++), JIT-compiled (Julia) or interpreted (Python). 
Especially notable in this sense is the Python API, EDIpack2py, which beyond providing a 
high-level interface for defining and solving impurity problems, also allows integration with established scientific libraries, such as NumPy and SciPy, facilitating data analysis and
rapid prototyping.

A key development feature in \NAME, unlocked by the introduction of the Python API, 
is the possibility to directly interface to the TRIQS (Toolbox for Research on Interacting Quantum Systems) and
W2Dynamics software suites, which are two of the most important and widely used computational tools in
the field. \NAME slots in as an alternative solver to the native Quantum Monte Carlo 
implementations, extending the parameter (and especially temperature) range reachable by 
the simulation while minimizing the coding requirements for the end user.
At the same time, this integration provides \NAME users with access to the
extensive data handling and analysis tools in the TRIQS ecosystem. 
Overall, interoperability extends the cumulative reach and trustworthiness of all these platforms,
enhancing reproducibility and cross-validation of numerical results, with the ultimate goal of achieving standardization for quantum impurity solvers along the guiding principles of initiatives such as FAIRmat, see \href{https://www.fairmat-nfdi.eu/fairmat/}{fairmat-nfdi.eu/fairmat}.

%\paragraph{closure}
In this paper, we present the mathematical framework, implementation
details, and key features of \NAME \textcolor{red}{5.0}, along with several illustrative
applications. We aim to demonstrate how this flexible,
high-performance solver can be used to tackle a wide range of
challenging impurity problems. 
The rest of this work has the following structure. In \secu{SecInstall} we 
briefly illustrate the overall structure of \NAME together with its dependencies, configuration
and installation procedures. In the following  \secu{SecEDIpack} we present the quantum impurity
problem and review the software implementation in full detail. In \secu{SecInterop} we thoroughly discuss the C-binding interface which is at the heart of the 
interoperability capabilities of \NAME. In this section we also discuss 4 interface layers: i) 
the Python API, ii) the TRIQS interface, iii) the W2Dynamics interface and iv) a basic Julia API.
Finally, in \secu{SecExamples} we discuss the functionalities and the capabilities of \NAME and 
all the discussed extensions with a series of illustrative examples.

### 2025-05-28 02_install.tex
Huge iteration
- Igor passed through a major part of the manuscript
- Samuele edited some parts of the e-ph section (response to Massimo)
- Adriano added author contributions
- Gabriele answered some comments by Igor and Lorenzo on the iRDM

--------

Co-authored-by: Igor Krivenko <iskrivenko@proton.me>
Co-authored-by: Lorenzo Crippa <crippa.lorenzo@gmail.com>
Co-authored-by: Samuele Giuli <samuele.giuli@gmail.com>
Co-authored-by: Adriano Amaricci <adriano.amaricci@gmail.com>

\section{Installation}\label{SecInstall}
The configuration and installation of \NAME is handled by CMake, which ensures
multi-platforms compabitility and dependencies resolution.  
The software builds into two distinct libraries.
The main one is a static Fortran library called {\tt  libedipack.a} which, alongside the compiled
modules, wraps the \NAME software in both its single- and multi-impurity variants.
A second dynamic library, {\tt libedipack\_cbindings.so}, together with the associated header
file {\tt edipack\_cbindings.h}, enables interoperability with other programming languages.

@@

\item{\bf EDIpack2ineq.}
  This extension, developed using modern Fortran constructs,
  enables the treatment of multiple, independent quantum 
impurities. It is particularly useful in DMFT applications involving 
unit cells with inequivalent atomic sites or systems with broken 
translational symmetry, such as heterostructures, large supercells, 
or disordered materials. This module provides flexible memory 
management and supports the simultaneous solution of multiple impurity 
problems, as discussed in Sec.\ref{sSecIneq}.

@@

\subsection{Dependencies}\label{sSecInstallDependencies}
\NAME directly depends on two external libraries.
\begin{itemize}
\item {\bf SciFortran}: an open-source Fortran library to support
  mathematical and scientific software development. 
\item {\bf MPI} (optional): a distributed memory parallel communication layer with support to modern Fortran compiler.
\end{itemize}

SciFortran provides a solid development platform enabling access to
many algorithms and functions, including standard linear algebra
operations and high-performance Lanczos based algorithms. This
greatly reduces code clutter and development time.
The use of distributed memory parallel environment, although optional,
is required to access scalable parallel diagonalization algorithms
which speed up calculations for large dimensional systems.

\subsection{Build and Install}\label{sSecInstallBuildInstall}
\subsubsection{Compilation from source}
The software can be installed from source as follows. The source code can
be retrieved directly from its GitHub repository, for instance using:
\begin{lstlisting}[style=mybash,numbers=none]
git clone https://github.com/edipack/EDIpack 
\end{lstlisting}
Then, assuming to be in the software directory, a conventional
out-of-source building can be performed using two alternative compilation backends.

@@

An alternative workflow employs the Ninja building backend with
Fortran support. Ninja is generally faster and automatically supports
multi-threaded building:
\begin{lstlisting}[style=mybash,numbers=none]
mkdir build
cd build
cmake -GNinja ..
ninja
ninja install
\end{lstlisting}
\end{itemize}

\noindent
The CMake configuration can be further tuned using the following variables:
\begin{center}
\begin{tabular}{ l|l|l } 
 \hline
  {\bf Option}               & {\bf Scope} & {\bf Value} \\
  \hline
  -D{\bf PREFIX}          & Install directory  & $\sim$/opt/EDIpack/TAG/PLAT/BRANCH\\
  -D{\bf USE\_MPI}       & MPI support  &  True/{\color{red}False}\\
  -D{\bf WITH\_INEQ}   & Inequivalent impurities support & {\color{red}True}/{False}\\
  -D{\bf VERBOSE}      & Verbose CMake output & {\color{red}True}/{False}\\ 
  -D{\bf BUILD\_TYPE} & Compilation flags & {\color{red}RELEASE}/TESTING/DEBUG/AGGRESSIVE \\
 \hline
\end{tabular}
\end{center}

@@

\subsubsection{Anaconda}
As an alternative, we provide for both Linux and macOS systems
installation through Anaconda packages into a virtual
environment containing Python ($>3.10$).

The Conda package installation procedure for a virtual environment called {\tt myenv} reads:
\begin{lstlisting}[style=mybash,numbers=none]
conda create -n myenv
conda activate myenv
conda install -c conda-forge -c edipack edipack
\end{lstlisting}
\noindent
which installs a bundle of SciFortran and \NAME libraries together with
specific {\tt pkg-config} configurations files, which can be used to
retrieve compilation and linking flags. In order to compile Fortran/C++ 
drivers, the {\tt compilers} conda package will need to be installed.

\subsection{OS Loading}\label{sSecInstallOSloading}
In order to avoid possible conflicts or the requirement for administrative
privileges, the building step results are installed by default in the user's {\tt HOME}
directory, specified by the CMake variable {\tt PREFIX}.
As a consequence, the system library and include paths will need to be updated by the user.

We offer different ways to perform this action:
\begin{enumerate}
\item  A CMake-generated configuration file for an environment module
  which allows to load and unload the library at any time. This is
  preferred solution for HPC systems. 
\item A CMake-generated bash script to be sourced (once or
  permanently) in any shell session to add \NAME library to the
  default environment.
\item A CMake generated {\tt pkg-config} file to be added in
  the {\tt PKG\_CONFIG\_PATH} itself.  
\end{enumerate}
An automatically generated recap message with all instructions is
generated at the end of the installation procedure.

>>>>

\section{Installation}\label{SecInstall}
The configuration and installation of \NAME is handled by CMake, which ensures
multi-platform compatibility and dependencies resolution.  
The software builds into two distinct libraries.
The main one is a static Fortran library called {\tt libedipack.a} which, alongside the compiled
modules, wraps the \NAME software in both its single- and multi-impurity variants.
A second dynamic library, {\tt libedipack\_cbindings.so}, together with the associated header
file {\tt edipack\_cbindings.h}, enables interoperability with other programming languages.

@@

\item{\bf EDIpack2ineq.}
  This extension, developed using suitable Fortran interfaces,
  enables the treatment of multiple, independent quantum 
impurities. It is particularly useful in DMFT applications involving 
unit cells with inequivalent atomic sites or systems with broken 
translational symmetry, such as heterostructures, large supercells, 
or disordered materials. This module provides flexible memory 
management and supports the simultaneous solution of multiple impurity 
problems, as discussed in \secu{sSecIneq}.

@@

\subsection{Dependencies}\label{sSecInstallDependencies}
\NAME directly depends on two external libraries.
\begin{itemize}
\item {\bf SciFortran}: an open-source Fortran library that provides support
for mathematical and scientific software development.
\item {\bf MPI} (optional): a distributed memory parallel communication layer with support for modern Fortran compiler.
\end{itemize}

SciFortran provides a solid development platform enabling access to
many algorithms and functions, including standard linear algebra
operations and high-performance Lanczos-based algorithms. This
greatly reduces code clutter and development time.
The use of distributed memory parallel environment, although optional,
is required to access scalable parallel diagonalization algorithms,
which speed up calculations for large systems.

\subsection{Build and Install}\label{sSecInstallBuildInstall}
\subsubsection{Compilation from source}
The software can be installed from source as follows. The source code can
be retrieved directly from its GitHub repository, for instance using
\begin{lstlisting}[style=mybash,numbers=none]
git clone https://github.com/edipack/EDIpack 
\end{lstlisting}
Then, assuming being in the software source directory, a conventional
out-of-source building can be performed using two alternative compilation backends.

@@

An alternative workflow employs the Ninja building backend with
Fortran support. Ninja is generally faster and automatically supports
parallel compilation:
\begin{lstlisting}[style=mybash,numbers=none]
mkdir build
cd build
cmake -GNinja ..
ninja
ninja install
\end{lstlisting}
\end{itemize}

\noindent
The CMake configuration can be further tuned using the following variables:
\begin{center}
\begin{tabular}{ l|l|l } 
 \hline
  {\bf Option}               & {\bf Scope} & {\bf Value} \\
  \hline
  -D{\bf PREFIX}          & Install directory  & {\color{red} $\sim$/opt/EDIpack/TAG/PLAT/BRANCH} \\
  -D{\bf USE\_MPI}       & MPI support  &  True/{\color{red}False}\\
  -D{\bf WITH\_INEQ}   & multi-impurities support & {\color{red}True}/{False}\\
  -D{\bf VERBOSE}      & Verbose CMake output & {\color{red}True}/{False}\\ 
  -D{\bf BUILD\_TYPE} & Compilation flags & {\color{red}RELEASE}/TESTING/DEBUG \\
 \hline
\end{tabular}
\end{center}

@@

\subsubsection{Anaconda}
As an alternative, we provide for both Linux and macOS systems
installation through Anaconda packages into a virtual
environment containing Python ($\geq3.10$).

The Conda package installation procedure for a virtual environment called {\tt myenv} reads
\begin{lstlisting}[style=mybash,numbers=none]
conda create -n myenv
conda activate myenv
conda install -c conda-forge -c edipack edipack
\end{lstlisting}
\noindent
which installs a bundle of SciFortran and \NAME libraries together with
specific {\tt pkg-config} configuration files, which can be used to
retrieve compilation and linking flags. In order to compile Fortran/C++ 
drivers, the {\tt compilers} conda package will need to be installed.

\subsection{Environment Variables}\label{sSecInstallOSloading}
In order to avoid possible conflicts or the requirement for administrative
privileges, the results of the building step are installed by default in user's {\tt HOME}
directory, specified by the CMake variable {\tt PREFIX}.
As a consequence, the environment variables holding the library and include paths will need to be updated by the user.
We offer different ways to perform this action:
\begin{enumerate}
\item  A CMake-generated configuration file for an environment module
  which allows to load and unload the library at any time. This is
  preferred solution for HPC systems. 
\item A CMake-generated bash script to be sourced (once or
  permanently) in any shell session to add \NAME library to the
  default environment.
\item A CMake generated {\tt pkg-config} file to be added in
  the {\tt PKG\_CONFIG\_PATH} itself.  
\end{enumerate}
An automatically generated recap message with all instructions is
generated at the end of the installation procedure.
